### 2023-07-11 calculus.tex
Updates from Overleaf

The marked lambda calculus is based on the gradually typed lambda calculus \cite{Siek06a} extended
with numbers, booleans, and empty expression holes \cite{HazelnutSNAPL}. The syntax is given in
\cref{fig:calculus-syntax}. There are two expression languages:
\begin{itemize}
  \item The \textbf{unmarked language}, which is our original language. Expressions of this language
    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.

  \item The \textbf{marked language}, which mirrors the structure of the unmarked language but is
    extended with explicit \emph{error marks}. We call expressions of this language
    \emph{marked expressions}, denoted $\ECMV$.
\end{itemize}
In this simple setting, we only need one sort for types, $\tau$. 
The base types $\TNum$ and $\TBool$ classify number and boolean
expressions. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$,
and there is a single arithmetic operation, addition. $\ETrue$ and
$\EFalse$ are the boolean values and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ is the boolean conditional. Arrow types classify lambdas in the usual way.

@@

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \mid \TProd{\TMV}{\TMV} \\
    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \mid \ELet{x}{\EMV}{\EMV}
                       \mid           \ENumMV \mid \EPlus{\EMV}{\EMV} \\
             &       & \mid         & \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
                       \mid           \EPair{\EMV}{\EMV}
                       \mid           \EEHole \\
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV}
                       \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV} \\
             &       & \mid         & \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV}
                       \mid           \ECEHole \\
             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV} \mid \ECInconAsc{\ECMV} \\
             &       & \mid         & \ECSynNonMatchedArrow{\ECMV} \mid \ECAnaNonMatchedArrow{\ECMV}
  \end{array}\]
  \caption{Syntax of the marked lambda calculus.}
  \label{fig:calculus-syntax}
\end{figure}

@@

The key operation is marking, which transforms an unmarked expression into a marked expression,
    inserting error marks where appropriate.
%
This corresponds to a type checking
process with error localization and recovery. The marked language may then serve as a foundation for
other semantic services, such as constraint-based inference (as discussed in
\cref{sec:thi}). 
Intuitively, each of the mark forms corresponds to a different kind of error message that might be shown by an editor or emitted by a compiler. (We do not specify what the error message should say in this paper.)

Throughout the remainder of this section, as we formulate marking for the GTLC, we will motivate and
give precise semantics for each error mark. When the core is extended, new marks may be needed, as we discuss below.

@@

Finally, the key judgement, \emph{marking}, is also specified bidirectionally. The synthetic marking judgment
$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
$\TMV$. Analogously, the analytic marking judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
states that $\EMV$ is marked into $\ECMV$, which analyzes against $\TMV$.

@@

Furthermore, since error marks are effectively annotations on top of the program, marking should
preserve syntactic structure modulo those marks. \cref{fig:calculus-mark-erasure} defines \emph{mark erasure}, which converts marked expressions back into unmarked expression by
removing error marks. 
\input{figs/calculus-mark-erasure}

Then, \cref{thm:calculus-marking-well-formedness} provides a
\emph{well-formedness} criterion for marking.
%
\begin{theorem}[name=Marking Well-Formedness] \
  \label{thm:calculus-marking-well-formedness}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then 
      $\erasesTo{\ECMV}{\EMV}$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then  and
      $\erasesTo{\ECMV}{\EMV}$.
  \end{enumerate}
\end{theorem}
%

These metatheorems imply that to go from a standard bidirectional system for the unmarked language to a marking system, we need to handle all possible failure modes with appropriate marks and marking logic (otherwise totality would be violated) and without otherwise changing the program (otherwise well-formedness would be violated).
Let us consider each form in turn.

@@

To start, let us consider the simple case of numbers. Because of \emph{subsumption}, which is
discussed in more detail next, we need only define a synthesis rule for unmarked numbers, in which
number literals synthesize the type $\TNum$:
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynTypeU{\ctx}{\ENumMV}{\TNum}
  }{USNum}
\end{mathpar}

How should numbers be marked? Quite straightforwardly, we simply give the same number as a marked
expression, synthesizing again $\TNum$. No type errors may occur in a single number, so we only need
the following rules for marking numbers and typing the results: 
%
\begin{mathpar}
  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\ENumMV}{\ECNumMV}{\TNum}
  }{MKSNum}

@@

Only synthetic rules were necessary for typing numbers and addition because of \emph{subsumption},
given below, which states that if an expression synthesizes a type, it may also be analyzed against
that type or any \emph{consistent} type.
\[%
  \judgment{
    \ctxSynTypeU{\ctx}{\EMV}{\TMV'} \\
    \consistent{\TMV}{\TMV'} \\
    \subsumable{\EMV}
  }{
    \ctxAnaTypeU{\ctx}{\EMV}{\TMV}
  }{UASubsume}
\]%
We rely on the notion of \emph{type consistency} from gradual type theory, which defines a reflexive
and symmetric (but not transitive) relation between types, writing $\consistent{\TMV_1}{\TMV_2}$ to
mean that $\TMV_1$ is consistent with $\TMV_2$. Defined in \cref{fig:calculus-consistency}, this
replaces the notion of \emph{type equality} to relate the unknown type to all other types.

@@

We also restrict the usage of subsumption to ``subsumable'' syntactic forms, written
$\subsumable{\EMV}$. This judgment is defined on all syntactic forms except lambda abstractions and
conditionals, the only ones with both synthesis and analysis rules (see below). In other words, we restrict subsumption to
be the rule of ``last resort'', which is necessary to establish that marking and typing are
deterministic (discussed later in \cref{thm:calculus-marking-unicity}).

@@

But what happens when the synthesized type of an expression is \emph{not} consistent with the
expected type, i.e. that the premise $\consistent{\TMV}{\TMV'}$ fails? Recalling the example of
\cref{fig:calculus-examples-inconsistent-types}, subsumption would be used to analyze $\ETrue$
against $\TNum$ in $\EPlus{\ETrue}{1}$, which fails since $\inconsistent{\TNum}{\TBool}$, i.e.
\textsc{UASubsume} does not apply, and traditional typing semantics would simply fail the type
checking process at this point. \textsc{MKASubsume} would also not be applicable, leaving marking
undefined in those cases.

@@

  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV}{\TMV'} \\
    \inconsistent{\TMV}{\TMV'} \\
    \subsumable{\ECMV}
  }{
    \ctxAnaTypeM{\ctx}{\ECInconType{\ECMV}}{\TMV}
  }{MAInconsistentTypes}
\end{mathpar}
%
Observe that the premises of \textsc{MKAInconsistentTypes} are identical to those of
\textsc{MKASubsume}, except that $\inconsistent{\TMV}{\TMV'}$. By marking an error, type-checking
may continue---instead of simply failing.

@@

However, consider the case that a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
bound. Again, \textsc{USVar} would not apply. A total marking should, however, report the error and
continue, motivating an \emph{unbound variable mark} $\ECUnbound{x}$ and the accompanying rules:
%
\begin{mathpar}
  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynFixedInto{\ctx}{x}{\ECUnbound{x}}{\TUnknown}
  }{MKSUnbound}

  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynTypeM{\ctx}{\ECUnbound{x}}{\TUnknown}
  }{MSUnbound}
\end{mathpar}
%
An unbound variable is marked as such, and since nothing may be said about their types, we
synthesize the unknown type. Similarly to the inconsistent type mark, this allows type checking to
proceed, with usage of the unbound variable permitted in any expression.

@@

The synthesis rule for marking intuitively follows \textsc{USLam} closely. We should recursively
mark the body with an extended context binding $x$ to $\TMV$ and construct a new marked lambda
abstraction:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\ELam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{MKSLam}

@@

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaTypeM{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxAnaTypeM{\ctx}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MALam1}
\end{mathpar}
%
However, recalling totality, we look again to the premises of \textsc{UALam} to see what type errors
may arise. First, what if $\TMV_3$ is not a matched arrow type? This would be the case if it were
$\TNum$, for example. The lambda abstraction's synthesized type would indeed be inconsistent with
the expected type, but this case is slightly distinct from that of the inconsistent types mark: the
lambda expression is in analytic position. Instead, we mark $\ECAnaNonMatchedArrow{\ECMV}$ to
indicate that an expression of the non-matched arrow type $\TMV_3$ was expected, but a lambda
abstraction was encountered.
%
\begin{mathpar}
  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TUnknown}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLamAnaNonMatchedArrow{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam2}

@@

In this example language, the distinction from the inconsistent type mark is of reduced
significance---all expressions synthesize a type. However, the addition of unannotated lambda
abstractions, for example, would necessitate such a disctinction.

Another error arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
$\inconsistent{\TMV}{\TMV_1}$, i.e. the actual type annotation for $x$ is inconsistent with the
expected input type. The \emph{inconsistent ascription mark} $\ECInconAsc{x}$ indicates exactly this
error, and we add a final pair of analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \inconsistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLamInconAsc{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam3}

@@

In the unmarked language, only a synthesis rule is necessary for applications:
\[%
  \judgment{
    \ctxSynTypeU{\ctx}{\EMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaTypeU{\ctx}{\EMV_2}{\TMV_1}
  }{
    \ctxSynTypeU{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\TMV_2}
  }{USAp}
\]%
Following the same methodology, we have the following marking and typing
rules:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
  }{MKSAp1}

@@

Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type, such
as in the example of \cref{fig:calculus-examples-app-non-lambda}. There is no expected type for the
argument, so we perform analytic marking on $\EMV_2$ against the unknown type. In this case, it is
not quite right to mark $\EMV_1$ with the inconsistent type mark; rather than any single type, it is
any type within the family of arrow types that is expected. For such a ``constrained'' synthetic
mode, we use a specialized mark $\ECSynNonMatchedArrow{\ECMV}$, which indicates that $\ECMV$ was
expected to be a function---but was not. The output type is unknown, so the entire expression
synthesizes the unknown type.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApSynNonMatchedArrow{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{MKSAp2}

@@

It is natural to extend this approach to other elimination forms that require the handling of
unmatched types, such as products. Indeed, the same approach is taken for projections in
\cref{sec:calculus-products}. 
% Another approach might indicate the same kind of error but mark the
% entire application.

@@

Conditionals, however, are an interesting case. In the unmarked language, we have both explicit
synthetic and analytic rules:
%
\begin{mathpar}
  \judgment{
    \ctxAnaTypeU{\ctx}{\EMV_1}{\TBool} \\\\
    \ctxSynTypeU{\ctx}{\EMV_2}{\TMV_1} \\
    \ctxSynTypeU{\ctx}{\EMV_3}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
  }{USIf}

@@

The framework also allows for other localization approaches. For example, as discussed at the
beginning of the section, we may regard the first branch as ``correct'' and localize any errors to
the second. The following rule formalizes such a design:
\[%
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
  }{MKSIf'}
\]%
% Of course, corresponding unmarked and marked typing rules would follow similar structure, and other
% heuristics may be devised for different kinds of branching structures.

>>>>

The marked lambda calculus is based on the gradually typed lambda calculus \cite{Siek06a} extended
with numbers, booleans, pairs, and empty expression holes \cite{HazelnutSNAPL}. Given in
\cref{fig:calculus-syntax}, the syntax consists of two expression languages:
\begin{itemize}
  \item The \textbf{unmarked language}, which is the original language. Expressions of this language
    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.

  \item The \textbf{marked language}, which mirrors the structure of the unmarked language but is
    extended with explicit \emph{error marks}. We call expressions of this language
    \emph{marked expressions}, denoted $\ECMV$.
\end{itemize}
In this simple setting, we only need one sort for types, $\tau$. 
The base types $\TNum$ and $\TBool$ classify number and boolean
expressions. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$,
and there is a single arithmetic operation, addition. $\ETrue$ and
$\EFalse$ are the boolean values and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ is the boolean conditional. Arrow and product types classify lambda abstractions and pairs, respectively, in the usual way.

@@

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
      \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \mid \TProd{\TMV}{\TMV} \\
      \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \mid \ELet{x}{\EMV}{\EMV}
                     \mid           \ENumMV \mid \EPlus{\EMV}{\EMV} \\
           &       & \mid         & \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
                     \mid           \EPair{\EMV}{\EMV}
                     \mid           \EProjL{\EMV} \mid \EProjR{\EMV}
                     \mid           \EEHole \\
      \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV}
                     \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV} \\
           &       & \mid         & \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV}
                     \mid           \ECPair{\ECMV}{\ECMV} \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
                     \mid           \ECEHole \\
           &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \\
           &       & \mid         & \ECLamInconAsc{x}{\TMV}{\ECMV} \mid \ECLamAnaNonMatchedArrow{x}{\TMV}{\ECMV} \mid \ECApSynNonMatchedArrow{\ECMV}{\ECMV} \\
           &       & \mid         & \ECInconBr{\ECMV}{\ECMV}{\ECMV} \\
           &       & \mid         & \ECPairAnaNonMatchedProd{\ECMV}{\ECMV} \mid \ECProjLSynNonMatchedProd{\ECMV} \mid \ECProjRSynNonMatchedProd{\ECMV}
  \end{array}\]
  \caption{Syntax of the marked lambda calculus.}
  \label{fig:calculus-syntax}
\end{figure}

@@

The key operation is \emph{marking}, which transforms an unmarked expression into a marked expression,
inserting error marks where appropriate.
%
This corresponds to a type checking
process with error localization and recovery. The marked language may then serve as a foundation for
other semantic services, such as constraint-based inference (as discussed in
\cref{sec:thi}). 
Intuitively, each of the mark forms corresponds to a different kind of error message that might be shown by an editor or emitted by a compiler. Note that we do not specify what the error message should say in this paper.
Throughout the remainder of this section, as we formulate marking for the GTLC, we will motivate and
give precise semantics for each error mark. Furthermore, when the core is extended, new marks may be needed, as we discuss below.

@@

Finally, the key judgement, \emph{marking}, is also of a bidirectional nature. The synthetic marking judgment
$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
$\TMV$. Analogously, the analytic marking judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
states that $\EMV$ is marked into $\ECMV$, which analyzes against $\TMV$.

@@

Furthermore, since error marks are effectively annotations on top of the program, marking should
preserve syntactic structure modulo those marks. \cref{fig:calculus-mark-erasure} gives selected clauses of the definition of \emph{mark erasure}, which converts marked expressions back into unmarked ones by
removing error marks. 
\input{figs/calculus-mark-erasure}

Then, \cref{thm:calculus-marking-well-formedness} provides the necessary
\emph{well-formedness} criterion for marking.
%
\begin{theorem}[name=Marking Well-Formedness] \
  \label{thm:calculus-marking-well-formedness}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then 
      $\erasesTo{\ECMV}{\EMV}$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then  and
      $\erasesTo{\ECMV}{\EMV}$.
  \end{enumerate}
\end{theorem}
%

Together, these metatheorems imply that to go from a standard bidirectional system for the unmarked language to a marking system, we need to handle all possible failure modes with appropriate marks and marking logic (otherwise totality would be violated) and without otherwise changing the program (otherwise well-formedness would be violated).
Now, let us consider each form in turn.

@@

To start, consider the simple case of numbers. Because of \emph{subsumption}, which is
discussed in more detail next, we need only define a synthesis rule for unmarked numbers, in which
number literals synthesize the type $\TNum$:
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynTypeU{\ctx}{\ENumMV}{\TNum}
  }{USNum}
\end{mathpar}

How should numbers be marked? Straightforwardly, we simply give the same number as a marked
expression, synthesizing again $\TNum$. No type errors may occur in a just single number, so we only need
the following rules for marking numbers and typing the results: 
%
\begin{mathpar}
  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\ENumMV}{\ECNumMV}{\TNum}
  }{MKSNum}

@@

Only synthetic rules are necessary for typing numbers and addition because of \emph{subsumption},
given below, which states that if an expression synthesizes a type, it may also be analyzed against
that type or any consistent type.
\[%
  \judgment{
    \ctxSynTypeU{\ctx}{\EMV}{\TMV'} \\
    \consistent{\TMV}{\TMV'} \\
    \subsumable{\EMV}
  }{
    \ctxAnaTypeU{\ctx}{\EMV}{\TMV}
  }{UASubsume}
\]%
We rely on the notion of \emph{type consistency} from gradual type theory, which defines a reflexive
and symmetric (but not transitive) relation between types, writing $\consistent{\TMV_1}{\TMV_2}$ to
mean that $\TMV_1$ is consistent with $\TMV_2$. Defined in \cref{fig:calculus-consistency}, this
replaces the notion of \emph{type equality} to relate the unknown type to all other types.

@@

We also restrict the usage of subsumption to ``subsumable'' syntactic forms, written
$\subsumable{\EMV}$. This judgment is defined for all syntactic forms except lambda abstractions,
conditionals, and pairs, the only ones with both synthesis and analysis rules (see below). In other words, we restrict subsumption to
be the rule of ``last resort'', which is necessary to establish that marking and typing are
deterministic (see \cref{thm:calculus-marking-unicity}).

@@

But what happens when the synthesized type of an expression is \emph{not} consistent with the
expected type, i.e. that the premise $\consistent{\TMV}{\TMV'}$ fails? Recalling the example of
\cref{fig:calculus-examples-inconsistent-types}, subsumption would be used to analyze $\ETrue$
against $\TNum$ when checking $\EPlus{\ETrue}{1}$, which fails since $\inconsistent{\TNum}{\TBool}$, i.e.
\textsc{UASubsume} does not apply. At this point, traditional typing semantics would simply fail the type
checking process. \textsc{MKASubsume} would also not be applicable, leaving marking
undefined in those cases.

@@

  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV}{\TMV'} \\
    \inconsistent{\TMV}{\TMV'} \\
    \subsumable{\ECMV}
  }{
    \ctxAnaTypeM{\ctx}{\ECInconType{\ECMV}}{\TMV}
  }{MAInconsistentTypes}
\end{mathpar}
%
Observe that the premises of \textsc{MKAInconsistentTypes} are identical to those of
\textsc{MKASubsume}, except that $\inconsistent{\TMV}{\TMV'}$. By marking an error, the type checking process may carry on and provide semantic feedback for the rest of the program.

@@

However, consider the case that a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
bound. Similar to above, \textsc{USVar} would not apply. A total marking procedure should, however, report the error and
continue, motivating an \emph{unbound variable mark} $\ECUnbound{x}$ and the accompanying rules:
%
\begin{mathpar}
  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynFixedInto{\ctx}{x}{\ECUnbound{x}}{\TUnknown}
  }{MKSUnbound}

  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynTypeM{\ctx}{\ECUnbound{x}}{\TUnknown}
  }{MSUnbound}
\end{mathpar}
%
An unbound variable is marked as such, and since nothing may be said about their types, we
synthesize the unknown type. As with the inconsistent type mark, this allows type checking to
proceed, with usage of the unbound variable permitted in any expression.

@@

The synthesis rule for marking intuitively follows \textsc{USLam} closely. We recursively
mark the body with an extended context and construct a new marked lambda
abstraction:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\ELam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{MKSLam}

@@

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaTypeM{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxAnaTypeM{\ctx}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MALam1}
\end{mathpar}

However, recalling totality, we look again to the premises of \textsc{UALam} to see what type errors
may arise. First, what if $\TMV_3$ is not a matched arrow type? This would be the case if it were
$\TNum$, for example. The lambda abstraction's synthesized type would indeed be inconsistent with
the expected type, but this case is slightly distinct from that of the inconsistent types mark: the
lambda expression is in analytic position. Instead, we mark $\ECAnaNonMatchedArrow{\ECMV}$ to
indicate that an expression of the non-matched arrow type $\TMV_3$ was expected, but a lambda
abstraction was encountered.
%
\begin{mathpar}
  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TUnknown}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLamAnaNonMatchedArrow{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam2}

@@

Note that in this example language, the distinction from the inconsistent type mark is of reduced
significance---all expressions synthesize a type. However, the addition of unannotated lambda
abstractions, for example, would necessitate such a disctinction.

Another error arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
$\inconsistent{\TMV}{\TMV_1}$, i.e. the actual type annotation for $x$ is inconsistent with the
expected input type. The \emph{inconsistent ascription mark} $\ECLamInconAsc{x}{\TMV}{\ECMV}$ indicates exactly this
error, and we add a final pair of analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \inconsistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLamInconAsc{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam3}

@@

In the unmarked language, only a synthesis rule is necessary for applications:
\[%
  \judgment{
    \ctxSynTypeU{\ctx}{\EMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaTypeU{\ctx}{\EMV_2}{\TMV_1}
  }{
    \ctxSynTypeU{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\TMV_2}
  }{USAp}
\]%
Following the same methodology up to this point, we have the following marking and typing
rules:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
  }{MKSAp1}

@@

Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type, such
as in the example of \cref{fig:calculus-examples-app-non-lambda}. There is no expected type for the
argument, so we perform analytic marking on $\EMV_2$ against the unknown type. In this case, it is
not quite right to mark $\EMV_1$ with the inconsistent type mark; rather than any single type, it is
any member of the family of arrow types that is expected. For such a ``constrained'' synthetic
mode, we use a specialized mark $\ECSynNonMatchedArrow{\ECMV}$, which indicates that $\ECMV$ was
expected to be a function---but was not. The output type is unknown, so the entire expression
synthesizes the unknown type.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
    \notMatchedArrow{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
  }{
    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApSynNonMatchedArrow{\ECMV_1}{\ECMV_2}}{\TUnknown}
  }{MKSAp2}

@@

It is natural to extend this approach to other elimination forms that require the handling of
unmatched types, such as products. Indeed, the same approach is taken for projections below.
Another similar approach might indicate the same kind of error but mark the entire application.

@@

Conditionals, however, present a more interesting case. In the unmarked language, we have both explicit
synthetic and analytic rules:
%
\begin{mathpar}
  \judgment{
    \ctxAnaTypeU{\ctx}{\EMV_1}{\TBool} \\\\
    \ctxSynTypeU{\ctx}{\EMV_2}{\TMV_1} \\
    \ctxSynTypeU{\ctx}{\EMV_3}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
  }{USIf}

@@

As previously mentioned, we do not prescribe any single localization design as ``correct'', and the framework freely allows for other approaches.
For example, as discussed in the introduction, we may choose to regard the first branch as ``correct'' and localize any 
errors to the second. The following rule formalizes such a design:
\[%
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV} \\
    \ctxAnaFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
  }{MKSIf'}
\]%
% Of course, corresponding unmarked and marked typing rules would follow similar structure, and other
% heuristics may be devised for different kinds of branching structures.

\subsubsection{Pairs}
At this point, the introduction of pairs expressions and the necessary projection operators poses no great challenge.
Explicit synthesis and analysis rules govern the former.
%
\begin{mathpar}
  \inferrule[USPair]{
    \ctxSynTypeU{\ctx}{\EMV_1}{\TMV_1} \\
    \ctxSynTypeU{\ctx}{\EMV_2}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TProd{\TMV_1}{\TMV_2}}
  }

  \inferrule[UAPair]{
    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaTypeU{\ctx}{\EMV_1}{\TMV_1} \\
    \ctxAnaTypeU{\ctx}{\EMV_2}{\TMV_2}
  }{
    \ctxAnaTypeU{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TMV}
  }
\end{mathpar}
%
In similar fashion to above, we may derive marking rules from these in an intuitive manner:
%
\begin{mathpar}
  \inferrule[MKSPair]{
    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\
    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
  }{
    \ctxSynFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TProd{\TMV_1}{\TMV_2}}
  }

  \inferrule[MSPair]{
    \ctxSynTypeM{\ctx}{\ECMV_1}{\TMV_1} \\
    \ctxSynTypeM{\ctx}{\ECMV_2}{\TMV_2}
  }{
    \ctxSynTypeM{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TProd{\TMV_1}{\TMV_2}}
  } \\
